---
title: "Statistical companion"
author: "Stefan Sobernig"
date: "`r Sys.Date()`"
output:
   html_document:
     toc: true
     dev: svg	  
   github_document:
     toc: true
     dev: png
---

Prerequisites
=========================

## How to (re-)generate the companion report

(This assumes that the R package `rmarkdown` is installed, in doubt,
run from within R: `install.packages("rmarkdown")`.)

1. Clone the repo or download its content
2. `cd replication/data`
3. `Rscript -e "rmarkdown::render('companion.Rmd','all')"`
4. Open `companion.html`

## R packages
```{r prereq}
wants <- c("ggplot2", "reshape2", "car", "pander")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has], repos = "http://cran.us.r-project.org")
```

```{r helpers, include=FALSE}

QQplot <- function(y, x = rnorm(sum(!is.na(y)))) {
  ## normalize the theoretical distribution data to scores ...
  ## x <- (x-min(x))/(max(x)-min(x))

  ## sort the data in ascending order
  d <- data.frame(x = sort(x), y = sort(y))
  p <- ggplot(d, aes(x=x, y=y)) + 
                  geom_point() + 
                      geom_smooth(method="lm", se = FALSE) + theme_bw() + 
                          xlab("Theoretical quantiles (normalized)") + 
                              ylab("Empirical quantiles")
  return(p)
}

my.nestedBoxplot <- function(data, response.var = "value") {
		 
    means <- melt(acast(data, DIST ~ FINT,
                        fun.aggregate = mean, value.var = response.var, na.rm = TRUE))
    
    se <- melt(acast(data, DIST ~ FINT,
                     function(x) { sd(x, na.rm = TRUE)/sqrt(length(x[!is.na(x)])) }, value.var = response.var))

    
    means$Var1 <- as.character(means$Var1)
    means$se <- se[,3]
    colnames(means) <- c("DIST","FINT","value","se")

ggplot(data) + 
    geom_boxplot(aes(x = factor(FINT), y = value, colour = DIST)) + 
    geom_point(data = means, aes(x = factor(FINT), y = value, shape=DIST)) +
    geom_line(data = means, aes(x = factor(FINT), y = value, group = DIST, linetype = DIST)) + 
    theme_bw(base_size = 16) + theme(panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank(), legend.justification=c(1,0), legend.position=c(1,0)) + ylab(unique(data$VARIABLE)[1]) + xlab("FINT")

}

my.interactionPlot <- function(data , response.var = "value") {
    means <- melt(acast(data, DIST ~ FINT,
                        fun.aggregate = mean, value.var = response.var, na.rm = TRUE))
    
    se <- melt(acast(data, DIST ~ FINT,
                     function(x) { sd(x, na.rm = TRUE)/sqrt(length(x[!is.na(x)])) }, value.var = response.var))

    
    means$Var1 <- as.character(means$Var1)
    means$se <- se[,3]

    ggplot(means, aes(x = factor(Var2), y = value, group = Var1)) + 
        geom_line(aes(linetype = Var1)) + geom_point(aes(shape = Var1)) +
            geom_errorbar(aes(ymax=value+se, ymin=value-se), width=.1) +
                theme_bw()
}

```


```{r data}
library(reshape2)
library(ggplot2)
library(car)
library(pander)

all <- read.csv2("toybox.csv")
levels(all$DIST) <- c("normal", "uniform", "x264", "origSayyadASE13")

df.toybox <- droplevels(subset(all, DIST != "origSayyadASE13"))
df.ase13 <- droplevels(subset(all, DIST == "origSayyadASE13"))
```

Data description
=========================

### ALGO=IBEA

```{r}
DATA <- subset(df.toybox, ALGO == "IBEA" & FINT %in% c("F","FI20","FI100"))
```	  

```{r, include=TRUE}
acast(DATA, DIST ~ FINT, length)
```

```{r}
ggplot(na.omit(subset(DATA, VARIABLE=="HV")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap(DIST ~ FINT, ncol = 3, drop = TRUE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("HV")

ggplot(na.omit(subset(DATA, VARIABLE=="PCORRECT")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap(DIST ~ FINT, ncol = 3, drop = TRUE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("PCORRECT")

ggplot(na.omit(subset(DATA, VARIABLE=="TimeToAnyC")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap(DIST ~ FINT, ncol = 3, drop = TRUE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("TimeToAnyC")
```

### ALGO=NSGAII

```{r}
DATA <- subset(df.toybox, ALGO == "NSGAII" & FINT %in% c("F","FI20","FI100"))
```	  

```{r, include=TRUE}
acast(DATA, DIST ~ FINT, length)
```


```{r}
ggplot(na.omit(subset(DATA, VARIABLE=="HV")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap(DIST ~ FINT, ncol = 3, drop = TRUE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("HV")

ggplot(na.omit(subset(DATA, VARIABLE=="PCORRECT" & value < 1000)), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap(DIST ~ FINT, ncol = 3, drop = TRUE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("PCORRECT")

ggplot(na.omit(subset(DATA, VARIABLE=="TimeToAnyC" & value < 50000)), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap(DIST ~ FINT, ncol = 3, drop = TRUE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("TimeToAnyC")
```

Analysis of variances (for ALGO = IBEA)
=========================

```{r}
DATA <- subset(df.toybox, ALGO == "IBEA" & FINT %in% c("F","FI100") & DIST %in% c("normal","x264"))
d.hv.glm <- glm(value ~ FINT * DIST, family = gaussian, data = subset(DATA, VARIABLE == "HV"))
d.pc.glm <- glm(value ~ FINT * DIST, family = gaussian, data = subset(DATA, VARIABLE == "PCORRECT"))
```

### Checks

#### Normality of residuals

```{r}
hv.res <- residuals(d.hv.glm)
QQplot(hv.res)

pc.res <- residuals(d.pc.glm)
QQplot(pc.res)
```

#### Homogeneity of variances


```{r}
hv <- subset(DATA, VARIABLE == "HV")
hv$combn <- interaction(hv$FINT,hv$DIST)
pc <- subset(DATA, VARIABLE == "PCORRECT")
pc$combn <- interaction(pc$FINT,pc$DIST)

ggplot(data=hv, aes(y = value, x = 1)) + geom_boxplot() + facet_wrap(~ combn, nrow=1) + theme_bw()
ggplot(data=pc, aes(y = value, x = 1)) + geom_boxplot() + facet_wrap(~ combn, nrow=1) + theme_bw()

```

```{r, results='asis'}

l.test <- cbind(as.numeric(leveneTest(value ~ FINT * DIST, data = hv)[1,]),
      as.numeric(leveneTest(value ~ FINT * DIST, data = hv, center = mean)[1,]),
                as.numeric(leveneTest(value ~ FINT * DIST, data = pc)[1,]),
      as.numeric(leveneTest(value ~ FINT * DIST, data = pc, center = mean)[1,]))

colnames(l.test) <- c("median","mean","median","mean")
rownames(l.test) <- c("Df", "F", "p-value")
pander(l.test)
```

### Anova table: Hypervolume (HV)

```{r}
ggplot(na.omit(subset(DATA, VARIABLE == "HV")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap(DIST ~ FINT, ncol = 2, drop = TRUE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("HV")
```

```{r, results='asis'}
panderOptions('digits', 4)
## panderOptions('round', 4)
panderOptions('keep.trailing.zeros', TRUE)
pander(anova(d.hv.glm, test = "F"))
```

```{r, results='asis'}
pander(aov(value ~ FINT * DIST, data = droplevels(subset(DATA, VARIABLE == "HV"))))
```

```{r}
my.interactionPlot(subset(DATA, VARIABLE == "HV"))
```

```{r hv, fig.height=4, fig.width=4}
my.nestedBoxplot(subset(DATA, VARIABLE == "HV"))
```

### Anova table: PCORRECT

```{r}
d.pc.glm <- glm(value ~ FINT * DIST, family = gaussian, data = subset(DATA, VARIABLE == "PCORRECT"))

ggplot(na.omit(subset(DATA, VARIABLE == "PCORRECT")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap(DIST ~ FINT, ncol = 2, drop = TRUE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("PCORRECT")
```

```{r, results='asis'}
pander(anova(d.pc.glm, test = "F"))
```


```{r, results='asis'}
pander(aov(value ~ FINT * DIST, data = droplevels(subset(DATA, VARIABLE == "PCORRECT"))))
```


```{r}
my.interactionPlot(subset(DATA, VARIABLE == "PCORRECT"))
```

```{r pc, fig.height=4, fig.width=4}
my.nestedBoxplot(subset(DATA, VARIABLE == "PCORRECT"))
```

### Significant differences: Tukey HSD

```{r, results="asis"}

hsd.pc <- TukeyHSD(aov(value ~ FINT * DIST, data = droplevels(subset(DATA, VARIABLE == "HV"))))
pander(hsd.pc$`FINT:DIST`)

hsd.pc <- TukeyHSD(aov(value ~ FINT * DIST, data = droplevels(subset(DATA, VARIABLE == "PCORRECT"))))
pander(hsd.pc$`FINT:DIST`)
```

Exact Replication
=========================

```{r, include=TRUE}
acast(df.ase13, VARIABLE ~ ALGO, length)
```

```{r}
ggplot(na.omit(subset(df.ase13, VARIABLE=="HV")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap( ~ ALGO, ncol = 2, drop = FALSE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("HV")

ggplot(na.omit(subset(df.ase13, VARIABLE=="PCORRECT")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap( ~ ALGO, ncol = 2, drop = FALSE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("PCORRECT")

ggplot(na.omit(subset(df.ase13, VARIABLE=="TimeTo50C")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap( ~ ALGO, ncol = 2, drop = FALSE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("TimeTo50C")


ggplot(na.omit(subset(df.ase13, VARIABLE=="TimeToAnyC")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap( ~ ALGO, ncol = 2, drop = FALSE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("TimeToAnyC")

```

```{r, results='asis'}
tmp.mean <- dcast(na.omit(subset(df.ase13, VARIABLE != "TimeTo50C")), VARIABLE ~ ALGO, mean)
tmp.sd <- dcast(na.omit(subset(df.ase13, VARIABLE != "TimeTo50C")), VARIABLE ~ ALGO, sd)
tmp.median <- dcast(na.omit(subset(df.ase13, VARIABLE != "TimeTo50C")), VARIABLE ~ ALGO, median)
tmp.mad <- dcast(na.omit(subset(df.ase13, VARIABLE != "TimeTo50C")), VARIABLE ~ ALGO, mad)
tmp.min <- suppressWarnings(dcast(na.omit(subset(df.ase13, VARIABLE != "TimeTo50C")), VARIABLE ~ ALGO, min))
tmp.max <- suppressWarnings(dcast(na.omit(subset(df.ase13, VARIABLE != "TimeTo50C")), VARIABLE ~ ALGO, max))
knitr::kable(cbind(mean = tmp.mean, sd = tmp.sd[,2:3], median = tmp.median[,2:3], mad = tmp.mad[,2:3], min = tmp.min[,2:3], max = tmp.max[,2:3]), format = "markdown", digits=4)
```

Exact vs. adjusted replication
=========================

```{r, include=TRUE}
df.contr <- rbind(cbind(df.ase13, EXP="exact"),
	 cbind(subset(df.toybox, FINT == "F" & DIST == "normal"), EXP="adjusted"))
acast(df.contr, EXP ~ ALGO, length)
```

```{r, include=TRUE}
ggplot(na.omit(subset(df.contr, VARIABLE=="HV")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap(EXP ~ ALGO, ncol = 2, drop = FALSE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("HV")

ggplot(na.omit(subset(df.contr, VARIABLE=="PCORRECT")), aes(y=value, x = 1)) +
    geom_violin() + geom_boxplot(width = 0.2) +
        facet_wrap(EXP ~ ALGO, ncol = 2, drop = FALSE) +
            stat_summary(fun.y="median", geom="point") +
                stat_summary(fun.y="mean", geom="point", shape=3) + xlab("PCORRECT")
```



Detach R packages
=========================

```{r}
try(detach(package:ggplot2))
try(detach(package:reshape2))
try(detach(package:car))
try(detach(package:pander))
```
